# Opening Files {#sec-open-files}

Most projects start the same way: you have a spreadsheet, and you need R to read it. This chapter covers the most common case, a plain `.csv` file.

## Reading a file with code

```r
survey <- read.csv("data/survey_results.csv")
head(survey)
```

```{r}
df <- data.frame(
  ID = seq(1:10),
  age = rnorm(10, mean = 30, sd = 5), 
  weigth = rnorm(10, mean = 70, sd = 10),
  height = rnorm(10, mean = 170, sd = 15))
```



```{r}

head(df, n=5)
```

The arrow `<-` means "store this on the left." This line reads: *read the file, and save the result into an object called `survey`.* From now on, typing `survey` shows you the data, and `head(survey)` shows just the first six rows.

::: {.callout-tip}
Above you can see the first five rows of the example data frame `df`, containing four columns: `ID`, `age`, `weight`, and `height`. The `head()` function is useful for quickly inspecting the first few rows of a data frame.
::: 

## Reading a file with the Import Dataset button

RStudio also offers a point-and-click route, which quietly writes the same code above for you:

1. Click **File → Import Dataset → From Text (readr)**
2. Browse to your file
3. Preview the table, adjust options if needed, then click **Import**

Watch the console after clicking Import — RStudio pastes the exact `read.csv(...)` line it used. This is a good way to learn the code by watching the point-and-click path generate it.

## Common file types

| File extension | Function |
|---|---|
| `.csv` | `read.csv()` |
| `.xlsx` | `readxl::read_excel()` |
| `.txt` (tab-separated) | `read.delim()` |
| `.rds` (R's own format) | `readRDS()` |

::: {.callout-note}
## Posit Cloud
Uploading a file works a little differently there than on your own computer. See [Chapter 8](08-posit-cloud.qmd) for the exact steps.
:::
