# Your First Plot {#sec-plot}

Once your numbers are in a vector, one line of code turns them into a picture. The built-in `hist()` function is the fastest way to see the shape of your data — how the values are distributed.

```r
hist(ages,
     main = "Distribution of ages",
     xlab = "Age",
     col  = "#001D41")
```

The chart below illustrates the idea: it redraws a histogram-shaped distribution with a different number of bins, the same way `hist()` would if you changed the `breaks` argument.

::: {.plot-box}
<div class="bins">
<button data-bins="5">5 bins</button>
<button class="active" data-bins="8">8 bins</button>
<button data-bins="12">12 bins</button>
</div>
<svg id="histSvg" viewBox="0 0 560 240" style="width:100%; height:auto;"></svg>
:::

Every plot you make in RStudio or Posit Cloud appears in the **Plots** pane, where an **Export** button lets you save it as a PNG or PDF directly into your `output/` folder.

::: {.callout-tip}
## When you're ready to go further
Once `hist()` and `plot()` feel comfortable, the `ggplot2` package (part of `tidyverse`) is the natural next step — same idea, far more control over how your charts look.
:::

## Plotting the NutriHeart data with `ggplot2`

Let's take that next step right away, using the same `nutriheart_data` dataframe we built in @sec-functions. `ggplot2` builds a chart in **layers**, always starting from the same three ingredients:

1. **Data** — which dataframe to draw from.
2. **Aesthetics** (`aes()`) — which columns map to which parts of the chart (x-axis, y-axis, color, ...).
3. **Geometry** (`geom_*()`) — what *kind* of shape represents the data: bars, points, boxes, lines.

You build a plot by adding these pieces together with `+`, the same way you chained dplyr steps together with `|>`.

```r
install.packages("ggplot2")
library(ggplot2)
```

::: {.callout-note}
## Why we rebuild the dataset here
Each `.qmd` file in this book renders in its **own, independent R session** — variables created in @sec-functions do not automatically carry over into this chapter. In your own projects you would normally avoid copy-pasting code like this by keeping the dataset in one script and loading it (e.g. with `readr::read_csv()`) in every chapter that needs it. For now, we simply rebuild `nutriheart_data` with the exact same code as before, so this chapter can run entirely on its own.
:::

```{r}
#| eval: TRUE
#| message: FALSE
library(tibble)
library(dplyr)
library(ggplot2)

set.seed(123)

nutriheart_data <- tibble(
  patient_id = 1:200,
  enrollment_date = sample(
    seq(as.Date("2020-01-01"), as.Date("2023-12-31"), by = "day"),
    size = 200, replace = TRUE
  ),
  gender = sample(c("Male", "Female"), size = 200, replace = TRUE),
  age = sample(18:85, size = 200, replace = TRUE),
  bmi = round(runif(200, min = 18.0, max = 42.0), 1),
  diet_type = factor(
    sample(c("Mediterranean", "Western", "Vegan", "Keto"),
           size = 200, replace = TRUE, prob = c(0.40, 0.30, 0.15, 0.15))
  ),
  physical_activity_level = factor(
    sample(c("Low", "Moderate", "High"), size = 200, replace = TRUE,
           prob = c(0.30, 0.45, 0.25)),
    levels = c("Low", "Moderate", "High"), ordered = TRUE
  ),
  daily_calories = round(rnorm(200, mean = 2200, sd = 450)),
  sugar_intake_g = round(rnorm(200, mean = 55, sd = 20), 1),
  hypertension = sample(c(TRUE, FALSE), size = 200, replace = TRUE, prob = c(0.35, 0.65)),
  smoker = sample(c(TRUE, FALSE), size = 200, replace = TRUE, prob = c(0.20, 0.80)),
  clinical_notes = sample(
    c("Patient reports fatigue.", "Adheres strictly to diet.",
      "Sedentary lifestyle.", "No recent complaints."),
    size = 200, replace = TRUE
  )
)
```

This is the identical `tibble()` call from @sec-functions — same `set.seed(123)`, same arguments — so it reproduces the exact same 200 rows. From this point on in the chapter, `nutriheart_data` behaves exactly as it did before.

### Histogram — one continuous variable

Just like `hist()`, but built layer by layer. Let's look at the distribution of `bmi`.

```{r}
#| eval: TRUE
ggplot(nutriheart_data, aes(x = bmi)) +
  geom_histogram(binwidth = 2, fill = "#001D41", color = "white") +
  labs(title = "Distribution of BMI", x = "BMI (kg/m²)", y = "Number of participants")
```

- `ggplot(nutriheart_data, aes(x = bmi))` creates the empty "canvas": it tells R which dataframe to use, and that whatever we draw next should read its x-position from the `bmi` column. Nothing is plotted yet — this line alone would produce a blank grid.
- `+ geom_histogram(...)` adds the actual bars. `binwidth = 2` sets each bar to cover a 2-unit-wide range of BMI (the equivalent of `hist()`'s `breaks`); `fill` colors the inside of the bars, `color` colors their outline.
- `labs(title = , x = , y = )` sets readable labels instead of the raw column name `bmi` — always label your axes for a reader who has never seen your code.

### Bar chart — one categorical variable

For a column like `diet_type`, there is no "distribution" to draw — instead, we count how many participants fall into each category.

```{r}
#| eval: TRUE
ggplot(nutriheart_data, aes(x = diet_type)) +
  geom_bar(fill = "#3E8E7E") +
  labs(title = "Participants per diet type", x = "Diet type", y = "Count")
```

`geom_bar()` is different from `geom_histogram()` in one important way: it counts rows *for you*, one bar per unique category found in `x`. You never need to compute the counts yourself with `table()` first — `ggplot2` does it internally.

### Boxplot — one continuous variable across groups

Now we connect a continuous variable to a categorical one — exactly the comparison our research question asked for: does BMI differ by diet type?

```{r}
#| eval: TRUE
ggplot(nutriheart_data, aes(x = diet_type, y = bmi)) +
  geom_boxplot(fill = "#F2A541") +
  labs(title = "BMI by diet type", x = "Diet type", y = "BMI (kg/m²)")
```

Here `aes()` maps two columns at once: `diet_type` to the x-axis (one box per category) and `bmi` to the y-axis (what each box summarizes). Each box shows the median (middle line), the interquartile range (the box itself — the middle 50% of values), and any unusually high or low values as individual dots.

### Line chart — a trend over time

Histograms, bar charts, and boxplots all summarize the data we already have as columns. A line chart is different: it is almost always used to show a **trend**, and a trend needs an x-axis that is ordered — usually time. `nutriheart_data` doesn't have an "enrollments per month" column yet, because nobody measured that directly; we have to **derive** it ourselves from `enrollment_date`.

```{r}
#| eval: TRUE
enrollment_by_month <- nutriheart_data |>
  mutate(enrollment_month = format(enrollment_date, "%Y-%m")) |>
  group_by(enrollment_month) |>
  summarise(n_enrolled = n()) |>
  arrange(enrollment_month) |>
  mutate(cumulative_enrolled = cumsum(n_enrolled))

head(enrollment_by_month)
```

This chunk creates **two brand-new, derived variables** that did not exist in the original dataset — a very common step before plotting a trend:

- `mutate(enrollment_month = format(enrollment_date, "%Y-%m"))` builds `enrollment_month` out of `enrollment_date`. `format(x, "%Y-%m")` rewrites a `Date` as text showing only its year and month (e.g. `2021-07`), which collapses every day in a month down to the same label — exactly what we need to count *monthly* enrollments.
- `group_by(enrollment_month) |> summarise(n_enrolled = n())` reuses the pattern from @sec-functions: split the data by month, then count how many rows (participants) fall in each one.
- `arrange(enrollment_month)` puts the months back in chronological order — since `enrollment_month` is text, R would otherwise sort it alphabetically, which happens to work for `"YYYY-MM"` labels, but it's good practice to be explicit.
- `mutate(cumulative_enrolled = cumsum(n_enrolled))` adds a second derived variable: the **running total** of participants recruited so far. `cumsum()` ("cumulative sum") adds each value to the sum of all values before it — turning monthly counts into a classic recruitment curve.

Now we can plot both derived variables as lines:

```{r}
#| eval: TRUE
ggplot(enrollment_by_month, aes(x = enrollment_month, y = n_enrolled, group = 1)) +
  geom_line(color = "#001D41", linewidth = 1) +
  geom_point(color = "#001D41") +
  labs(title = "New participants enrolled per month",
       x = "Month", y = "Number enrolled") +
  theme(axis.text.x = element_text(angle = 90))
```

A couple of details are new here, and both matter every time you draw a line over time:

- `group = 1` tells `ggplot2` "treat every row as part of the *same single line*." Without it, `ggplot2` sees a text column on the x-axis and, not knowing how to connect separate categories together, refuses to draw a continuous line. This is a very common beginner error — if your line chart comes out empty, check for a missing `group`.
- `geom_line()` connects the points in the order of the x-axis; adding `geom_point()` on top just marks each actual monthly value, which makes the chart easier to read.
- `theme(axis.text.x = element_text(angle = 90))` rotates the month labels 90 degrees so they don't overlap each other — a small but frequent fix once your x-axis has many labels.

The same idea works for the cumulative version — only the `y` and the label change:

```{r}
#| eval: TRUE
ggplot(enrollment_by_month, aes(x = enrollment_month, y = cumulative_enrolled, group = 1)) +
  geom_line(color = "#3E8E7E", linewidth = 1) +
  labs(title = "Cumulative participant enrollment",
       x = "Month", y = "Total participants enrolled") +
  theme(axis.text.x = element_text(angle = 90))
```

This second chart is a good example of why deriving variables is worth the extra step: `cumsum()` turned a simple monthly count into a recruitment curve, a chart type you will see constantly in the methods section of real cohort studies.

### Scatter plot — two continuous variables, colored by a third

Finally, let's look at the relationship between `age` and `bmi`, and use color to bring in a fourth piece of information: whether the participant has `hypertension`.

```{r}
#| eval: TRUE
ggplot(nutriheart_data, aes(x = age, y = bmi, color = hypertension)) +
  geom_point(alpha = 0.6) +
  labs(title = "Age vs. BMI, colored by hypertension status",
       x = "Age (years)", y = "BMI (kg/m²)", color = "Hypertension")
```

- Adding `color = hypertension` inside `aes()` tells `ggplot2` to draw each point in a different color depending on that participant's `TRUE`/`FALSE` value — it automatically builds a legend for us.
- `geom_point()` draws one dot per row of the dataframe.
- `alpha = 0.6` makes each point slightly transparent (60% opaque), which helps when points overlap — a simple but important trick once you have 200 dots on the same chart.

::: {.callout-tip}
## The pattern to remember
Every `ggplot2` chart you will ever build follows the same skeleton:

```r
ggplot(your_data, aes(x = ..., y = ...)) +
  geom_something() +
  labs(title = "...", x = "...", y = "...")
```

Only the dataframe, the columns inside `aes()`, and the `geom_*()` you choose ever really change.
:::

## Function finder

Not sure which function you need? Choose a task below.

::: {.matcher-box}
<select id="matcherSelect2">
<option value="">Choose a task…</option>
<option value="hist">I want to see the shape of one numeric variable</option>
<option value="bar">I want to count how many rows fall into each category</option>
<option value="box">I want to compare a numeric variable across groups</option>
<option value="scatter">I want to see the relationship between two numeric variables</option>
<option value="line">I want to see how a value changes over time</option>
</select>
<pre class="matcher-out" id="matcherOut2"></pre>
:::

::: {.quiz-box}
**Quick check:** in `ggplot2`, which function actually draws the bars, points, or lines onto the canvas?

<div class="opt" data-correct="false"><code>aes()</code></div>
<div class="opt" data-correct="false"><code>ggplot()</code></div>
<div class="opt" data-correct="true"><code>geom_*()</code></div>
<div class="fb"></div>
:::

::: {.quiz-box}
**Quick check:** your `geom_line()` chart shows nothing but an empty grid. The x-axis is text (like `enrollment_month`). What did you most likely forget?

<div class="opt" data-correct="false"><code>labs()</code></div>
<div class="opt" data-correct="true"><code>group = 1</code> inside <code>aes()</code></div>
<div class="opt" data-correct="false"><code>library(ggplot2)</code></div>
<div class="fb"></div>
:::

::: {.quiz-box}
**Quick check:** to compare BMI across the four diet types in one chart, the best choice is...

<div class="opt" data-correct="false"><code>geom_histogram()</code></div>
<div class="opt" data-correct="true"><code>geom_boxplot()</code></div>
<div class="opt" data-correct="false"><code>geom_bar()</code></div>
<div class="fb"></div>
:::
